---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  # fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 7.2916667,
  fig.height = 7.2916667,
  dev = "svg"
)

library(unusualprofile)
library(simstandard)
library(ggnormalviolin)
library(dplyr)
```
# unusualprofile

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/ggnormalviolin)](https://cran.r-project.org/package=unusualprofile) [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/wjschne/unusualprofile.svg?branch=master)](https://travis-ci.org/wjschne/unusualprofile)
[![Codecov test coverage](https://codecov.io/gh/wjschne/unusualprofile/branch/master/graph/badge.svg)](https://codecov.io/gh/wjschne/unusualprofile?branch=master)
<!-- badges: end -->

The goal of unusualprofile is to calculate conditional Mahalanobis distances and related statistics. Such statistics can help find cases with unusual profiles of multivariate normal data.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wjschne/unusualprofile")
```
## Example

Using the simstandard package, we generate 1 case from a standardized multivariate normal data set. 

```{r data}
library(unusualprofile)
library(simstandard)
library(ggnormalviolin)
library(dplyr)
set.seed(281)

# lavaan model with three indicators of a latent variable
model <- "
X =~ 0.7 * X_1 + 0.5 * X_2 + 0.8 * X_3
Y =~ 0.7 * Y_1 + 0.5 * Y_2 + 0.8 * Y_3
Y ~ 0.6 * X
"

# Randomly generated case
d <- sim_standardized(
  model, 
  n = 1, 
  observed = TRUE, 
  latent = FALSE, 
  errors = FALSE, 
  composites = TRUE) %>% 
  rename(X = "X_Composite",
         Y = "Y_Composite")



# Model-implied correlation matrix
R <- sim_standardized_matrices(model)$Correlations$R_all
```


```{r, eval=FALSE, include=FALSE}
library(tidyverse)
find_un <- function(seed = 1) {
  set.seed(seed)
  d <- sim_standardized(
  model, 
  n = 1, 
  observed = TRUE, 
  latent = FALSE, 
  errors = FALSE, 
  composites = TRUE) %>% 
  rename(X = "X_Composite",
         Y = "Y_Composite")

  cm <- cond_maha(data = d, 
          R = R,
          v_ind = c("X_1", "X_2", "X_3"),
          v_dep = c("Y_1", "Y_2", "Y_3"))

tibble(p = cm$dCM_p, ind_mean = mean(d$X_1, d$X_2, d$X_3), id = seed)
  
}

map_df(200:1000, find_un) %>% 
  filter(p > 0.95, ind_mean > 1 | ind_mean < -1)
```


Now we specify the correlations, independent variables, and dependent variables. In this case, the indpendent variables are composite scores summarizing the dependent variables.

```{r example}
# Conditional Mahalanobis distance
cm <- cond_maha(data = d, 
          R = R,
          v_ind = c("X_1", "X_2", "X_3"),
          v_dep = c("Y_1", "Y_2", "Y_3"))

cm

# Plot
plot_cond_maha(cm)

# Conditional Mahalanobis distance
cm <- cond_maha(data = d, 
          R = R,
          v_ind = c("X", "Y"),
          v_dep = c("X_1", "X_2", "X_3",
                    "Y_1", "Y_2", "Y_3"))

cm

# Plot
plot_cond_maha(cm)

unusualprofile::unusualness(d, model, 
                            v_dep = c("Y"),
                            v_ind = c("X")) %>% 
  mutate(Measure = stringr::str_replace(Measure, "\\$\\\\leftarrow\\$","&larr;")) %>% 
  knitr::kable(digits = 2) 




```

