---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 7.2916667,
  fig.height = 7.2916667,
  dev = "svg"
)
options(digits = 2)
library(unusualprofile)
library(simstandard)
library(ggnormalviolin)
library(dplyr)
```
# unusualprofile

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/ggnormalviolin)](https://cran.r-project.org/package=unusualprofile) [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/wjschne/unusualprofile.svg?branch=master)](https://travis-ci.org/wjschne/unusualprofile)
[![Codecov test coverage](https://codecov.io/gh/wjschne/unusualprofile/branch/master/graph/badge.svg)](https://codecov.io/gh/wjschne/unusualprofile?branch=master)
<!-- badges: end -->

The goal of unusualprofile is to calculate conditional Mahalanobis distances and related statistics. Such statistics can help find cases with unusual profiles of multivariate normal data.

# Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("wjschne/unusualprofile")
```
# Example

To use the unusualprofile package, all that is needed is to know the correlations, means, and standard deviations among a set of continuous variables and at least one row of data from that set of variables. 

Suppose we have set of variables that have the following relationships:

![Multivariate normal model](man/figures/rm_model.svg)

First, we load the packages that we will use later:

```{r}
library(unusualprofile)
library(ggnormalviolin)
library(dplyr)
```


The case in the `d_example` data set was generated from the model depicted above:

```{r, echo=FALSE}
d_example
```

The model-implied correlation matrix:

```{r}
R_example
```

## Using the `cond_maha` function

Now we specify the correlations (`R`), means (`mu`), standard deviations (`sigma`). independent variables (`v_ind`), and dependent variables (`v_dep`). In this case, the indpendent variables are composite scores summarizing the dependent variables.

```{r example}
# Conditional Mahalanobis distance
cm <- cond_maha(data = d_example, 
          R = R_example,
          mu = 0,
          sigma = 1,
          v_ind_composites = c("X_Composite", "Y_Composite"),
          v_dep = c("X_1", "X_2", "X_3",
                    "Y_1", "Y_2", "Y_3"))

cm

# Plot
plot_cond_maha(cm)
```

